Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-11-14 pkg/circtree/inst/wind_paper/draft/draft.Rnw
revisions on motivational example and introduction to trees
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2875 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 03ce88331d21eef3a914f85febae014619c07f17

As a first illustration Figure~\ref{fig:tree_ibk} displays a distributional
tree applied to predict the wind direction at the airport of Innsbruck in the following hour. 
Innsbruck lies at the bottom of a deep valley in the Alps. Topography channels wind along the 
west-east valley axis or along a tributary valley intersecting from the south. Hence, pressure 
gradients to which valley wind regimes react both west and east of the airport are considered 
as covariates along with other meteorological measurements at the airport (lagged by one hour),
such as wind direction and wind speed at Innsbruck Airport.

Together with the tree structure itself Figure~\ref{fig:tree_ibk} also depicts the empirical (gray) 
and fitted von Mises (red) distribution of wind direction in each terminal node. Based on the 
fitted location parameters $\hat \mu$, the subgroups can be distinguished into the following 
wind regimes:
(1)~Up-valley winds blowing from the valley mouth towards the upper valley (from
east to west, nodes 4 and 5). 
(2) Downslope winds blowing across the Alpine crest along the intersecting valley towards 
Innsbruck (from south-east to north-west, nodes 7 and 8). 
(3) Down-valley winds blowing in the direction of the valley mouth (from west to east, 
nodes 12, 14 and 15). 
Node~11 captures observations with rather low wind speeds that cannot be distinguished
clearly into wind regimes and consequently are associated with a very low
estimated concentration $\hat \kappa$. In terms of covariates, the lagged
wind direction (``persistence'') is mostly responsible for distinguishing
the broad wind regimes listed above while the pressure gradients and wind
speed separate between subgroups with high vs.\ low precision.
Note that in the meteorological context wind direction is defined on the scale $[0,360]$ degree and 
increases clockwise from North ($0$ degree).

@@

The corresponding methodology and its features are presented in Section~\ref{sec:tree_forest} after
discussing distributional models for circular data in Section~\ref{sec:prob_circ}.
In Section~\ref{sec:wind} the presented models are applied to probabilistic forecasting of wind direction
at the airports of Innsbruck and Vienna.

@@

\section{Circular regression trees and forests}
\label{sec:tree_forest}
\fixme{revise transition from sec\,2. to sect\,3. with the focus on circular methods}
A common way of including covariates as regressors in distributional models is to apply
generalized additive models for location, scale, and shape 
\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005} where each distribution parameter is estimated
by its own generalized additive model (GAM).
As already discussed in Section~\ref{sec:introduction}, additive strategies work well 
for smooth effects but difficulties can arise in case of non-additive changes. Additionally, 
specifying a model properly can be very challenging, particularly for a high number of covariates 
and no information on possible interactions.
An alternative regression approach that can deal with these problems is to apply a tree-structured 
model.

>>>>

As a motivational example Figure~\ref{fig:tree_ibk} displays a distributional
tree applied to predict the wind direction at the airport of Innsbruck in the following hour. 
Innsbruck lies at the bottom of a deep valley in the Alps. Topography channels wind along the 
west-east valley axis or along a tributary valley intersecting from the south. Hence, pressure 
gradients to which valley wind regimes react both west and east of the airport are considered 
as covariates along with other meteorological measurements at the airport (lagged by one hour),
such as wind direction and wind speed at Innsbruck Airport.

Together with the tree structure itself Figure~\ref{fig:tree_ibk} also depicts the empirical (gray) 
and fitted von Mises (red) distribution of wind direction in each terminal node. Based on the 
fitted location parameters $\hat \mu$, the subgroups can be distinguished into the following 
wind regimes:
(1)~Up-valley winds blowing from the valley mouth towards the upper valley (from
east to west, nodes 4 and 5). 
(2) Downslope winds blowing across the Alpine crest along the intersecting valley towards 
Innsbruck (from south-east to north-west, node 8). 
(3) Down-valley winds blowing in the direction of the valley mouth (from west to east, 
nodes 10, 12 and 13). 
Node~7 captures observations with rather low wind speeds that cannot be distinguished
clearly into wind regimes and consequently are associated with a very low
estimated concentration $\hat \kappa$. In terms of covariates, the lagged
wind direction (``persistence'') is mostly responsible for distinguishing
the broad wind regimes listed above while the pressure gradients and wind
speed separate between subgroups with high vs.\ low precision.
Note that in the meteorological context wind direction is defined on the scale $[0,360]$ degree and 
increases clockwise from North ($0$ degree).

@@

A more extensive case study of circular regression trees and forests applied to probabilistic 
forecasting of wind direction at the airports of Innsbruck and Vienna is presented in 
Section~\ref{sec:wind} where their performance is also compared to other commonly used models.
Before that, distributional models for circular data are discussed in Section~\ref{sec:prob_circ} 
and the methodology of circular regression trees and forest and its features are presented in 
Section~\ref{sec:tree_forest}.

@@

\section{Circular regression trees and forests}
\label{sec:tree_forest}
\fixme{revise transition from sec\,2. to sect\,3. with the focus on circular methods}
A common way of including covariates is to regress the model parameters on the covariates.
In the case of a probabilistic model for a circular response this corresponds to expressing the 
location parameter $\mu$ and the concentration parameter $\kappa$ of the von Mises distribution
by means of the available covariates, e.g., in the form of a linear combination, possibly with an 
appropriate link function plugged in as in GLMs \citep{Fisher+Lee:1992, Fisher:1993}.
In a more general setting this idea can be extended by generalized additive models for location, scale, 
and shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005} where each distribution parameter is estimated
by its own generalized additive model (GAM). However, an adaption of this methodology to circular data
can not be obtained by the straight forward approach of simply wrapping the modelled additive effects
around the unit circle since the parameters of the von Mises distribution and their dependence/interaction 
differs strongly from the linear case and therefore requires specific handling.
Moreover, difficulties might arise in case of non-additive changes and specifying a model properly can 
be very challenging, particularly for a high number of covariates and no information on possible 
interactions.
An alternative regression approach that can deal with these problems is to apply a tree-structured 
model.
